---
title: 'Part 1: Exploring Exponential Distributions and thier Means'
author: "Kimberly Bormann"
date: "2025-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

## Overview


##Simulations


#Single Sample of Exponential Variables nd its Theortical Distribution
In this report we will explore two distribution. The first is a 1000 random exponentiation variables with λ = 0.2. Given our knowledge of exponential functions, we expect that when we visualize this distribution it will look very similar to the function used to generate the random variables which is `r $f(x) = 0.2e^{-0.2x}$`, where x is randomly generated.

First we generate the variables.

```{r}
        exp_raw <- rexp(1000, rate = lambda)
        cat("First few entries of a single simulation of 1000 values:\n")
        head(exp_raw)
```

Now, let's do the simple calculations for the theoretical means and variance for an exponential distribution with 1000 occurrences and λ = 0.2. 

```{r}
lambda <- 0.2
theoretical_mean <- 1/lambda
cat("The theoretical mean is", theoretical_mean, "\n")
theoretical_sd <- 1/lambda
theoretical_var <- (1/lambda)^2
cat("The theoretical variance is", theoretical_var, "\n")
```

Then we can view the distribution to see if it reflect out expectation that it is right skewed like the decreasing exponential function. To make this easier, I overlay the function `r $f(x) = 0.2e^{-0.2x}$` on the histogram.

```{r, fig.width=5, fig.height=5}
        #histogram form 1000 exponential values
         hist(expDis, probability = TRUE, 
             main = "Histogram of 1000 Exponential Values", 
             xlab = "Exponential Value", 
             col = "lightblue",
             xlim = c(0, quantile(expDis, 0.99)))
        
        #overlay the theoretical exponential density curve
        curve(dexp(x, rate = lambda), 
              from = 0, 
              to = quantile(expDis, 0.99),
              add = TRUE, 
              col = "red", 
              lwd = 2)
```

This density function is right-skewed. We can see that the random sample is a very close match to the theoretical density function because the Law of Large Numbers(LLN) tells us that the large the sample the more likely it is to approach the theoretical density curve. 

Lastly, let's compare the theoretical mean and variance to the sample mean and variance.

Mean of this single simulation:" `r mean(exp_raw)`
Variance of a sinlge simulation: `r var(exp_raw)`

We can see that these the sample and theoretical means and variance are very similar. 


#1000 Sample of 40 Exponential Variables and its Theortical Distribution

The theoretical means and variance for 1000 samples of 40 exponential random variables with λ = 0.2.
```{r}
#theoretical mean calulation
lambda <- 0.2
theoretical_mean <- 1/lambda
cat("The theoretical mean is", theoretical_mean, "\n")
theoretical_sd <- 1/lambda

#theoretical variance calulation
n <- 40
theoretical_sd <- 1/lambda
theoretical_se <- theoretical_sd/sqrt(n)
theoretical_mean_var <- theoretical_se^2
cat("The theoretical variance of the means is", theoretical_mean_var, "\n")
```

Next, let's generate the values and compare the mean and variance of hte sample to the theoretical mean and variance. 

```{r}
#generate 1000 simulation of 40 random exponential variables in a matrix where each column contains 40 random exponential variables.
n_sim <- 1000
sim_matrix <- matrix(rexp(n*n_sim, rate = lambda), nrow = n, ncol = n_sim)
#calculate the mean of each sample
sim_means <- apply(sim_matrix, 2, mean)
#calculate the variance of each sample
sim_vars <- apply(sim_matrix, 2, var)
#vector of all the sample means
avg_values <- replicate(1000, mean(rexp(n, rate = lambda)))
cat("first few sample means:\n")
head(sim_means)
```

Now let's compare the sample means and variance to the theoretical ones calulates earlier. 

```{r}
cat("Sample Means (averages of 40 exponentials):\n")
cat("  Sample Mean:", round(mean(sim_means), 3), "\n")
cat("  Sample Variance:", round(var(sim_means), 3), "\n\n")

cat("Theoretical Values:\n")
cat("  Theoretical Mean:", theoretical_mean, "\n")
cat("  Theoretical Variance of the Sample Mean:", round(theoretical_mean_var, 3), "\n")
```

As you can see the sample mean and variance is extremely close to the theoretical mean and variance. 

Lastly, let's view the distribution. Unlike the first simulation which generated 1000 random exponential variables, this simulation generated 1000 means of random exponential variables. 


Therefore, by the Cental Limit Theorem(CLT), we expect that this distribution will be normally distributed because CLT state that, although each individual observation is not normally distributed, the distribution of the sample mean will tend toward a normal distribution as the sample size increases. With sample size 1000 we expect a mostly normal distribution. 

For comparison,  I overlay the normal distribution with the theoretical mean and variance to the sample distribution. 

```{r fig.width=5, fig.height=5}
#plot the histogram of sample means with density scaling
        hist(sim_means, probability = TRUE,
             main = "Histogram of 1000 Sample Means (n = 40)",
             xlab = "Sample Mean",
             col = "lightgreen")
        
#overlay the theoretical normal density curve
        curve(dnorm(x, mean = theoretical_mean, sd = sqrt(theoretical_mean_var)),
              from = min(sim_means), to = max(sim_means),
              add = TRUE, col = "red", lwd = 2)        
```

We can see that the distribution is normally distributed as expected. 



